---
title: "Econ 294 Assignment 5"
author: "Curtis Kephart"
date: "Winter 2016"
output: pdf_document
---

This assignment is designed to get you to explore `ggplot2`.
 

Please refer to the ggplot2 documentation website for help. 
I also found UPENN's ggplot2 intro tutorial from 2010 useful. 

Due by Feb 26th 2016 (after the next lecture). Turn in your `.R` script by pushing it to your public github repo  
and emailing the URL to your instructor (at `curtisk+econ294_05@ucsc.edu`) 



```{r, echo = F, message = F, warning= F}
#load data
library(ggplot2)
library(dplyr)
library(foreign)

org_example <- read.dta("/Users/OKComputer/Dropbox/R Lab/econ294_2015/data/org_example.dta")
```


1. From ggplot2's diamond's dataset, approximately replicate the following plots. 

a. Scatter plot of diamond `prices`, volume (`x * y * z`), with point size `carat`, color `clarity`, and the x-axis and y-axis log  scale (e.g `scale_x_log10`). 

```{r, echo = F, message = F, warning= F, fig.height = 4}
ggplot(diamonds, 
       aes(x = x*y*z, y = price, size = carat, group = clarity, fill = clarity, colour = clarity)) +
  geom_point(alpha = 0.2) +
  scale_x_log10() +
  scale_y_log10() +
  guides(fill = guide_legend(override.aes= list(alpha = 0.9, 
                                                size = 6))) 
```


\pagebreak



b. Stacked histogrames showing density of diamonds by carat, stacked by clarity, broken down by cut type. (Check out `geom_histogram`, `facet_grid`, and the use of `..density..` in `geom_histogram`)
```{r, echo = F, message = F, warning= F, fig.height = 4.5}

	ggplot(diamonds, 
	       aes(carat, ..density..)) +
	    geom_histogram(aes(fill = clarity), binwidth = 0.2) +
	    facet_grid(cut ~ .) +
		labs(y='Density', x='Carat', fill='Clarity')
```


c. Violin plots of price, by cut. With jittered, very high transparency scatterplots of price. (`geom_violin`, `geom_jitter`, `alpha`. )

```{r, echo = F, message = F, warning= F, fig.height = 3.}

	ggplot(diamonds, 
	       aes(x = cut, y = price)) +
  geom_violin() +
  geom_jitter(alpha = 0.01)
```

\pagebreak



3. From the `org_example` dataset, approximately replicate the following plots. 


a. Plot monthly median `rw` in a line plot, along with the band for the first and third quartiles, and first and ninth deciles. (There may be many ways to do this, I used dplyr single table verbs, plus `geom_line`, `geom_ribbon`, `alpha` and `ylim`. )


```{r, echo = F, message = F, warning= F, fig.height = 2.5}

df <- org_example %>%
  mutate(
    date = paste(year, month, "01", sep = "-"),
    date = as.Date(date, format = "%Y-%m-%d")
  ) %>%
  filter(!is.na(rw)) %>%
  tbl_df() %>%
  group_by(date) %>%
  summarize(
    Median.RW = median(rw),
    RW.1stQr  = quantile(rw, 0.25),
    RW.3rdQr  = quantile(rw, 0.75),
    RW.9thDc  = quantile(rw, 0.90),
    RW.1stDc  = quantile(rw, 0.10)
  )

ggplot(
  df, 
  aes(x = date)
) + 
  geom_line(aes(y = Median.RW)) + 
  geom_ribbon(aes(ymin = RW.1stDc, ymax = RW.9thDc), alpha = 0.2)+
  geom_ribbon(aes(ymin = RW.1stQr, ymax = RW.3rdQr), alpha = 0.4) +
  ylim(0, 50)

```



b. Plot monthly median real wages in a line plot, broken down into education level.

```{r, echo = F, message = F, warning= F, fig.height = 2.5}
df <- org_example %>%
  mutate(
    date = paste(year, month, "01", sep = "-"),
    date = as.Date(date, format = "%Y-%m-%d")
  ) %>%
  filter(!is.na(rw)) %>%
  tbl_df() %>%
  group_by(date, educ) %>%
  summarize(
    Median.RW = median(rw)
  )

ggplot(
  df, 
  aes(x = date, colour = educ, fill = educ)
) + 
  geom_line(aes(y = Median.RW))

```





